Supermartingales and Termination Analysis of Statistical PCF

### 2020-09-28 probabilisticProgrammingMartingales.tex
Start writing out the full proof of the canonicalCousins lemma.

\begin{lemma} \label{canonicalCousins}
For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ (for each position in the final term in $X$) and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
\end{lemma}

@@

\begin{proof}
Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur. It is only required that the reductions that do occur occur in the correct order.
\lo{This paragraph (especially the last sentence) is ambiguous.
What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
\iffalse
\lo{So this is permitted?
\[
(II)(I (I \bar I)) \to
(II)(I\bar I) \to
(II)\bar I 
\]
but not this: 
\(
(II)(I (I \bar I)) \to
(II)(I \bar I) \to
(II)\bar I 
\)?}
\fi

Proof sketch:
Starting at $X$, take the last pair of adjacent reductions which aren't in CBV order, and switch them (with $\sim_c$). Repeat this process until CBV order is reached.

If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$.

If Y is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
\end{proof}

>>>>

\begin{lemma} \label{canonicalCousins}
For any reduction sequence $X$, there is a unique reduction sequence $X_{cbv}$ that is related to $X$ by $\sim_c$ and is in call-by-value order, and if $X \sim_c^* Y$, then $X_{cbv} = Y_{cbv}$.
\end{lemma}

@@

\begin{proof}
A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underscore f_i$ and $\underscore f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order. Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur. It is only required that the reductions that do occur occur in the correct order.
\lo{This paragraph (especially the last sentence) is ambiguous.
What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
\iffalse
\lo{So this is permitted?
\[
(II)(I (I \bar I)) \to
(II)(I\bar I) \to
(II)\bar I 
\]
but not this: 
\(
(II)(I (I \bar I)) \to
(II)(I \bar I) \to
(II)\bar I 
\)?}
\fi

Define $X_n$ recursively so that $X_0 = X$ and if $X_n$ is not in CBV order, take the last pair of reductions in $X_n$ which aren't in CBV order. They are adjacent in the sequence, as the CBV order is a total order. Let the positions of these reductions be $\alpha$ and $\beta$ respecively. If they are disjoint, they can be considered as one of the branches of $\sim_c$ case a. Otherwise, $\beta < \alpha$, and the sub-skeleton at position $\beta$ at the appropriate point in the reduction sequence is a redex with non-trivial subterms, therefore it's of one of the forms $\tif{\skeletonPlaceholder>0}{A}{B}$, $(\lambda x.A) B$ or $\tY \lambda x. A$, therefore the reductions $\alpha$ and $\beta$ form the $O_2$ branch of one of the cases b, c, d, e or f of $\sim_c$. In either case, define $X_{n+1}$ as equal to $X_n$ except taking the other (left) branch, so that the order of the reductions at $\alpha$ and $\beta$ are switched (although the equivalent(s) of the reduction at $\alpha$ may not actually be at that position).

This sequence of swaps rearranges the reductions in $X$ rather like insertion-sort (but in some cases duplicating or deleting the new element), and eventually reaches some $X_n$ that is in CBV order. This is defined to be $X_{cbv}$.

\paragraph{}
If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$. To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$. The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions. If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already. Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order). If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it. If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$. If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.

The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possible overlapping. If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order. This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$. As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.

In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$. There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$. Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps. Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the position of the first relevant reduction in $Y$ be $\delta$. Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ results in $\gamma, \beta_k, \dots, \beta_1$. In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.

\begin{itemize}
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a. Eventually it reaches some $X_i$ where the relevant sub-reduction-sequence is $\alpha$ then whatever $\beta_1$ was turned into in $X \sim_c X_1$ then $\gamma$, before moving on to swapping pairs earlier in the reduction sequence. The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above. The subskeleton at $\alpha$ at the relevant point in the reduction sequence if the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma$ is left of $\alpha$: In this case, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$, therefore $Y_1 = X$ already.
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma < \alpha$: Again, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$ therefore $Y_1 = X$.
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 1 (so that $\gamma = \beta_1$): In this case, $Y \sim_c X$ either by case b or by case e (where the lambda has no occurrences of its variable), therefore $\gamma < \delta$ (the position of the reduction which is essentially deleted in $Y \sim_c X$). $\alpha$ is either outside or left of $\gamma$, therefore it bears the same relation to $\delta$. The sequence $(X_i)_i$ starts by swapping $\gamma$ below $\alpha$, and $(Y_j)_j$ starts by swapping $\gamma$ then $\delta$ below $\alpha$, then what happens next depends on the shape of the term at $\alpha$, or equivalently, by which case of $\sim_c$ are $\gamma$ and $\delta$ swapped below $\alpha$. If it's cases a, c or d, each of $\gamma$ and $\delta$ corresponds to one reduction in $Y_2$, and the relative position of these is the same as the relative position of $\gamma$ and $\delta$ therefore the next step of $(X_i)_i$ involves these two reductions and deletes the one equivalent to $\delta$, so $Y_3 = X_1$. If it's case b or e, in $X_2$ there are $n$ reductions corresponding to $\delta$ and $n$ reductions corresponding to $\gamma$, and they are all disjoint except that for each image of $\delta$, there's one image of $\gamma$ that it's inside, in a position that is deleted by the reduction at that image of $\gamma$. The next $\frac{n(n+1)} 2$ steps in $(X_i)_i$ are then each image of $\delta$ in turn being swapped (by case a) past the images of $\gamma$ until they reach the one they match, at which point they get deleted. In this case therefore, $Y_{n(n+1)/2+2} = X_1$. Case f, corresponding to $\tY$-reduction, is more complicated because the images of $\gamma$ and $\delta$ are not all disjoint. For each of $\gamma$ and $\delta$, there is one image of it that may contain other such positions. Let these be $\gamma_o$ and $\delta_o$. The sequence $(Y_j)_j$ then proceeds by moving all of the images of $\delta$ which are right of $\gamma_o$ to their corresponding $\gamma$-image and deleting them, then moving all the images of $\delta$ which are after $\delta_o$ but which are inside $\gamma_o$ (one at a time) past all the images of $\gamma$ left of $\gamma_o$ by case a, then either deleting them against $\gamma_o$ by case b or e as appropriate, or moving them past it by case c or d, then possibly past some of the images of $\gamma$ inside $\gamma_o$, then deleting them once they reach the matching image of $\gamma$. Then $\delta_o$ itself is moved to $\gamma_o$ then deleted, then the remaining images of $\delta$ are similarly moved and deleted. The end result of this process is that all the images of $\delta$ are deleted and all the images of $\gamma$ are unaffected, therefore for some $j$, $Y_j = X_1$.
    \item \akr{There are several more cases to go, and that last case is more typical of how complex they'll be. I'd guess around 5 more, but I haven't worked them all out precisely. Also, I might find a neater way to deal with case f, or at least not have to repeat that argument, because the case where both reductions are $\tY$-reductions should include all the complexity of any other case involving case f.}
\end{itemize}

\paragraph{}
If Y is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
\end{proof}
